---
title: "Graph Customisation"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

# Introduction

When you first create plots in R, the default settings work well for exploring your data. However, when you're preparing visualizations for presentations, reports, or publications, you'll want to polish them to make your message clearer and more visually appealing.

In this tutorial, we'll learn how to customize various aspects of ggplot2 graphs including axes, colors, labels, legends, and themes. We'll primarily use the `palmerpenguins` dataset, which contains measurements of three penguin species from islands in Antarctica.

First, let's load our packages and data:

```{r}
#| message: false
library(ggplot2)
library(palmerpenguins)
library(dplyr)

# View the first few rows
head(penguins)
```

# Customizing Axes

Axes help your readers understand the scale and range of your data. Let's explore how to modify them.

## Numeric Axes

You can control numeric axes using `scale_x_continuous()` and `scale_y_continuous()`. The most useful options are:

- `breaks`: Where to place tick marks and labels
- `limits`: The minimum and maximum values to display

```{r}
#| warning: false
# Basic plot
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() +
  scale_x_continuous(breaks = seq(30, 60, by = 5),
                     limits = c(30, 60)) +
  scale_y_continuous(breaks = seq(13, 22, by = 2),
                     limits = c(13, 22)) +
  labs(title = "Penguin Bill Dimensions",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)")
```

**Tip**: The `seq()` function creates a sequence of numbers. For example, `seq(30, 60, by = 5)` creates: 30, 35, 40, 45, 50, 55, 60.

## Formatting Numbers

The `scales` package provides helpful functions for formatting axis labels:

- `comma`: Adds commas to large numbers (1000 becomes 1,000)
- `dollar`: Adds currency symbols
- `percent`: Converts decimals to percentages

```{r}
# Create sample data
set.seed(42)
population_data <- data.frame(
  city = paste("City", 1:40),
  population = rnorm(40, 250000, 80000),
  employment_rate = runif(40, 0.55, 0.85),
  median_income = rnorm(40, 65000, 15000)
)

ggplot(population_data, aes(x = population, y = employment_rate)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Population vs Employment Rate",
       x = "City Population",
       y = "Employment Rate")
```

## Categorical Axes

For categorical variables, use `scale_x_discrete()` or `scale_y_discrete()`:

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "coral") +
  scale_x_discrete(labels = c("Adélie\nPenguin", 
                              "Chinstrap\nPenguin", 
                              "Gentoo\nPenguin")) +
  labs(title = "Number of Penguins by Species",
       x = "Species",
       y = "Count")
```

**Note**: `\n` creates a line break in text labels.

## Date Axes

Working with dates requires `scale_x_date()` or `scale_y_date()`:

```{r}
# Using the economics dataset
ggplot(economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = "darkblue", linewidth = 1) +
  scale_x_date(date_breaks = "10 years",
               date_labels = "%Y") +
  labs(title = "US Unemployment Over Time",
       x = "Year",
       y = "Unemployed (thousands)")
```

Common date format codes:

| Code | Meaning | Example |
|------|---------|---------|
| %d | Day of month | 01-31 |
| %m | Month number | 01-12 |
| %b | Abbreviated month | Jan |
| %B | Full month | January |
| %y | 2-digit year | 24 |
| %Y | 4-digit year | 2024 |

# Working with Colors

Colors can make your plots more engaging and help distinguish between groups.

## Setting Colors Manually

Use `color` for points and lines, and `fill` for bars and areas:

```{r}
#| warning: false
# Single color
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "darkgreen", size = 2) +
  labs(title = "Penguin Bill Measurements")
```

To assign specific colors to categories, use `scale_color_manual()` or `scale_fill_manual()`:

```{r}
#| warning: false
ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_manual(values = c("Adelie" = "darkorange",
                                "Chinstrap" = "purple",
                                "Gentoo" = "cyan4")) +
  labs(title = "Penguin Species Count",
       x = "Species",
       y = "Count",
       fill = "Species")
```

## Using Color Palettes

Pre-designed color palettes often look more professional than colors you pick yourself.

### ColorBrewer Palettes

ColorBrewer provides carefully designed color schemes:

```{r}
#| warning: false
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Penguin Species by Island",
       x = "Species",
       y = "Count",
       fill = "Island")
```

### Viridis Palettes

Viridis palettes are colorblind-friendly and print well in grayscale:

```{r}
#| warning: false
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(title = "Penguin Species by Island",
       x = "Species",
       y = "Count",
       fill = "Island")
```

Use `_c` for continuous variables (numbers) and `_d` for discrete variables (categories).

# Customizing Points and Lines

## Point Shapes

Change point shapes with the `shape` parameter:

```{r}
#| warning: false
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                     shape = species)) +
  geom_point(size = 3) +
  labs(title = "Bill Dimensions by Species",
       shape = "Species")
```

Shapes 21-25 allow both fill and border colors:

```{r}
#| warning: false
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                     fill = species)) +
  geom_point(shape = 21, size = 3, color = "black") +
  labs(title = "Bill Dimensions by Species")
```

## Line Types

Modify line appearance with `linetype`:

```{r}
ggplot(economics, aes(x = date)) +
  geom_line(aes(y = unemploy), linetype = "solid", color = "blue") +
  geom_line(aes(y = unemploy * 1.1), linetype = "dashed", color = "red") +
  labs(title = "Unemployment Trends",
       y = "Number Unemployed")
```

# Labels and Legends

## Adding Labels

Use the `labs()` function to add or modify all text in your plot:

```{r}
#| warning: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, 
                     color = species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title = "Penguin Body Mass vs Flipper Length",
    subtitle = "Measurements from Palmer Station, Antarctica",
    caption = "Data source: palmerpenguins package",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Penguin Species"
  ) +
  theme_minimal()
```

## Legend Position

Control legend placement with `theme()`:

```{r}
#| warning: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, 
                     color = species)) +
  geom_point(size = 2) +
  labs(title = "Penguin Measurements",
       color = "Species") +
  theme_minimal() +
  theme(legend.position = "top")
```

Options for `legend.position`:
- `"top"`, `"bottom"`, `"left"`, `"right"`
- `c(x, y)` for exact placement (e.g., `c(0.9, 0.2)` for bottom-right inside plot)
- `"none"` to remove the legend

# Adding Annotations

Annotations highlight important features in your data.

## Text Annotations

Use `annotate()` to add text anywhere on your plot:

```{r}
#| warning: false
penguin_summary <- penguins %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "lightblue") +
  annotate("text", x = 2, y = 6000, 
           label = "Gentoo penguins are\nthe heaviest species",
           color = "darkred", size = 4) +
  labs(title = "Body Mass Distribution by Species",
       x = "Species",
       y = "Body Mass (g)")
```

## Adding Reference Lines

Use `geom_hline()` and `geom_vline()` for horizontal and vertical lines:

```{r}
#| warning: false
mean_mass <- mean(penguins$body_mass_g, na.rm = TRUE)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, 
                     color = species)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = mean_mass, 
             linetype = "dashed", color = "darkred") +
  annotate("text", x = 175, y = mean_mass + 200,
           label = "Overall Mean", color = "darkred") +
  labs(title = "Penguin Measurements with Mean Body Mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

## Highlighting Specific Groups

The `gghighlight` package makes it easy to emphasize particular data:

```{r}
#| warning: false
library(gghighlight)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(size = 2, color = "coral") +
  gghighlight(species == "Gentoo") +
  labs(title = "Highlighting Gentoo Penguins",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

# Using Themes

Themes control the overall appearance of your plot (backgrounds, gridlines, fonts, etc.).

## Built-in Themes

ggplot2 includes several pre-made themes:

```{r}
#| warning: false
p <- ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  labs(title = "Penguins by Species and Island")

# Try different themes
p + theme_minimal()  # Clean and simple
```

```{r}
#| warning: false
p + theme_classic()  # Traditional look
```

```{r}
#| warning: false
p + theme_dark()  # Dark background
```

## Custom Themes with ggthemes

The `ggthemes` package offers many professional themes:

```{r}
#| warning: false
library(ggthemes)

base_plot <- ggplot(penguins, aes(x = flipper_length_mm, 
                                   y = body_mass_g, 
                                   color = species)) +
  geom_point(size = 2) +
  labs(title = "Penguin Physical Measurements",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")

# Economist style
base_plot + theme_economist() + scale_color_economist()
```

```{r}
#| warning: false
# FiveThirtyEight style
base_plot + theme_fivethirtyeight()
```

# Combining Multiple Plots

The `patchwork` package lets you arrange multiple plots together:

```{r}
#| warning: false
#| message: false
library(patchwork)

# Create three plots
p1 <- ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(title = "Species Count") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(title = "Body Mass by Species",
       y = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, 
                            color = species)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  labs(title = "Flipper Length vs Body Mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)") +
  theme_minimal()

# Combine plots: | means side-by-side, / means stacked
(p1 | p2) / p3 +
  plot_annotation(
    title = "Palmer Penguins: A Multi-faceted View",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

The `&` operator applies settings to all plots:

```{r}
#| warning: false
(p1 | p2) / p3 & theme_light()
```

# Tips for Effective Visualizations

1. **Keep it simple**: Don't add decorations that don't help tell your story
2. **Label everything**: Make sure axes, legends, and titles are clear
3. **Consider your audience**: Choose colors and styles appropriate for your readers
4. **Test accessibility**: Ensure your plots work in grayscale and for colorblind readers
5. **Be consistent**: Use the same style across all plots in a report or presentation

# Summary

You've learned how to:

- Customize axes for numeric, categorical, and date data
- Apply colors using manual specification and palettes
- Modify point shapes and line types
- Add labels, legends, and annotations
- Apply and customize themes
- Combine multiple plots into one figure

These skills will help you create professional, publication-ready visualizations. Practice by customizing your own data visualizations!

# Further Resources

- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Data Visualization: A Practical Introduction](https://socviz.co/)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)